SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data

### 2011-08-31 seqbias-paper.tex
finishing touches

\begin{figure}
\centerline{\includegraphics[width=0.30\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the model is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

@@

To determine the structure and parameters of the Bayesian network, we use a
hill-climbing approach similar to the algorithm described by \citet{Grossman2004}.
The network structure is determined by greedily optimizing the conditional
log-likelihood:
\begin{align*}
\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) \\
&=
\sum_{i=1}^{n} \log \frac{ \Pr(s_i | x_i) \Pr( x_i ) }{
\sum_{x \in \{0,1\}} \Pr( s_i | x ) \Pr(x) } \\
\end{align*}
where $\Pr(x)$ is flat (i.e.  $\Pr( x = 0 ) = \Pr( x = 1 ) =
0.5$) if we sample foreground and background positions equally.

As we will be estimating parameters and evaluating the likelihood on the same
set of samples, simply maximizing the likelihood would severely overfit the
training set. We thus penalize model complexity heuristically using the Bayesian
Information Criterion \citep{Schwarz1978a}. Where $m$ is the number of
parameters needed to specify the model, we maximize,

$$ \ell' = 2 \ell - m \log n $$

@@

If the null model has log-likelihood $L$, and the bias-corrected model $L'$, a
simple goodness of fit measure is the improvement in log-likelihood (a statistic
commonly known as McFadden's pseudo-coefficient of determination
\citep{McFadden1974}), defined as,
$$R^2 = 1 - \frac{L'}{L}$$

@@

We have demonstrated that sequence bias can confound, sometimes severely,
quantification in RNA-Seq experiments, and we have introduced an effective
method to account for this bias without the need of existing gene annotations.

In our results, estimating initial heptamer frequencies was not seen to be as
effective as the other models, even when data generated using random hexamer
priming was used. Given the large number of parameters needed to estimate
heptamer frequencies, a likely explanation is that the model is overfit to the
training set. Though we trained the 7mer model on at least 600,000 reads in each
experiment, it seems more reads, or a greater diversity of reads are needed
to produce an accurate or useful estimate.

>>>>

\begin{figure}
\centerline{\includegraphics[width=0.25\textwidth]{overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is trained to discriminate between the set of
sampled foreground and background sequences, (d) and the model is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

@@

To determine the structure and parameters of the Bayesian network, we use a
hill-climbing approach similar to the algorithm described by \citet{Grossman2004}.
The network structure is determined by greedily optimizing the conditional
log-likelihood:
\begin{align*}
\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) 
=
\sum_{i=1}^{n} \log \frac{ \Pr(s_i | x_i) \Pr( x_i ) }{
\sum_{x \in \{0,1\}} \Pr( s_i | x ) \Pr(x) } \\
\end{align*}
where $\Pr(x)$ is flat (i.e.  $\Pr( x = 0 ) = \Pr( x = 1 ) =
0.5$) if we sample foreground and background positions equally.

As we will be estimating parameters and evaluating the likelihood on the same
set of samples, simply maximizing the likelihood would severely overfit the
training set. We thus penalize model complexity heuristically using the Bayesian
Information Criterion \citep{Schwarz1978a}. Where $m$ is the number of
parameters needed to specify the model, we maximize, 
$$ \ell' = 2 \ell - m \log n $$

@@

If the null model has log-likelihood $L$, and the bias-corrected model $L'$, a
simple goodness of fit measure is the improvement in log-likelihood (a statistic
commonly known as McFadden's pseudo-coefficient of determination
\citep{McFadden1974}), defined as,
$R^2 = 1 - L'/L$.

@@

We have demonstrated that sequence bias can confound, sometimes severely,
quantification in RNA-Seq experiments, and we have introduced an effective
method to account for this bias without the need of existing gene annotations.
The analysis provided demonstrates that our method shows significant improvement
in two aspects: uniformity of read coverage, and consistency of nucleotide
frequencies. These results strongly suggest an increase in accuracy, and we show
in Supplementary Section 8, that this conclusion is further justified by
comparison to qRT-PCR.

In our results, estimating initial heptamer frequencies was not seen to be as
effective as the other models, even when data generated using random hexamer
priming was used. Given the large number of parameters needed to estimate
heptamer frequencies, a likely explanation is that the model is overfit to the
training set.  We trained the 7mer model on a minimum of 600,000 reads (with the
Bullard data set), and up to 23 million reads (with the Wetterbom data set), yet
it seems more reads, or a greater diversity of reads are needed to produce an
accurate or useful estimate.

### 2011-08-31 seqbias-supplement.tex
finishing touches

>>>>

\section{Comparison to qRT-PCR}

We used sequencing data previously published by Au, et al, \cite{Au2010} to
evalute the effect bias correction has on correlation to measurements made by
TaqMan RT-PCR, made available by the Microarray Quality Control project
\cite{Shi2006}. The RNA-Seq data shows a pattern of bias similar to that seen in
other samples sequenced on an Illumina platform, as plotted below.

\centerline{\includegraphics[width=0.6\textwidth]{fig/au-freqs.pdf}}

To evaluate the efficacy of each of the bias correction methods considered,

We counted reads overlapping each gene, defining the gene by the union of every
transcript annotated in release 62 of the Ensembl gene annotations, excluding
UTRs, and 100 additional bases from both ends of the gene. Counts were then
normalized (divided) by the adjusted length of these genes.  We then remove any
genes with reads fewer than 10 reads, or without a unique TaqMan probe, leaving
592 genes.

Each bias correction method was trained using a procedure identical to the one
described in the results section. After adjusting read counts according to the
predicted sequence bias, we compute the Pearson correlation between log read counts
and log TaqMan expression values, which are averaged across three replicates. The
expression values

These correlations are listed below.

\begin{center}
\begin{tabular}{ll}
\textbf{Method} & \textbf{Correlation} \\ \hline
Unadjusted &  0.6665 \\
7mer &  0.6677 \\
GLM & 0.6854 \\
MART & 0.6921 \\
BN & 0.7016
\end{tabular}
\end{center}

These results are consistent with with the KL-divergence and Poisson regression
analysis presented in Section 3. We see only very small gains with the
brute-force 7mer method. The GLM and MART models show much better performance,
and our Baysian network (BN) method shows moderate increase correlation over these.
